### 2023-12-22 thesis/MSECE_thesis.tex
Final edits from committee.

\usepackage{cite}
\usepackage{listings}
\usepackage{multicol}
\usepackage{newtxtext}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\usepackage[T1]{fontenc}
% \usepackage[numbib]{tocbibind}
\usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}

@@

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades,
achieving performance levels good enough to be brought to consumer-facing products such as personal assistants, search engines, phones,
televisions, etc. Due to the success in the general domain, applications have begun to spin off into problem-specific domains, such as
(and most notably for this work) aviation. For example, several speech recognition systems with rule- and machine learning-based
natural language processing, inference, and semantic extraction algorithms have been implemented at EUROCONTROL simulation centers in
Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM) systems on
controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large
natural language corpora such as WikiText \cite{merity_pointer_2016} and IMDB \cite{maas_learning_2011}. In contrast to the general domain, the
aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications. Typically, the most
prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of
the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for
aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}, making it significantly more difficult to
transcribe/label data as it requires transcribers to have domain knowledge and experience to create effective transcriptions. This, in addition to the
following facets, further contributes to the difficulty of transcription: (1) the medium of communication, (2) the environment in which the
communication occurs\footnote{For example, a noisy and possibly low fidelity medium, depending on the equipment used to transmit and receive
    communications}, (3) a noisy and active communication environment i.e.~an active flight deck under the influence of wind, engine noise, etc., (4)
an active Air Traffic Control (ATC) tower and channels, and (5) a rapid rate of speech \cite{paltridge_handbook_2013}.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since
2019\footnote{https://rajpurkar.github.io/SQuAD-explorer/}) \cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
aviation domain. If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems,
they could easily boost efficiency, reduce workloads, and reduce and mitigate errors during normal operations.

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep
learning-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the
language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
results show that language models trained ``from scratch'' on in-domain data perform better than models trained on out-of-domain data and transfer
learned on in-domain data, i.e., constrained BERT and RoBERTa models transfer learned to aviation English.

\section{Notation and Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.

@@

\noindent
as the union of the set $A$ and a set containing the element $x$. The equation below can be read as adding the element $x$ to the set $A$:
\begin{equation*}
    A \leftarrow A; x
\end{equation*}

The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:
\begin{itemize}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence (distinct from characters and words)
    \item The rules for the formation of tokens from a sentence (and into sequences) shall be defined by tokenization algorithms (Section
          \ref{sec:tokenizers})
    \item The definition of \textbf{words} and \textbf{sentences} follow from that of the English language
    \item The \textbf{vocabulary}, \textbf{inventory}, \textbf{index}, and \textbf{word list} of a model shall be semantically identical
\end{itemize}

\section{Literature Review}\label{sec:lit_review}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art
\cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations and understandings of languages to
predict token probabilities instead of transforming one language into another \cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been
found to be very effective at other NLP-related tasks such as prompt completion and sentiment analysis among others (i.e.~auto-regressive and sequence
classification tasks, respectively) \cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods
such as masked language modeling and next sentence prediction to improve natural language understanding for downstream tasks, which was further
expanded upon by Liu et al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

% tokenizer algorithms
In language modeling and natural language processing tasks, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by the
introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding (BPE) have been used to solve OOV problems while simultaneously minimizing
vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.

% LMs and NLP in aero. domain
Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications and try to mitigate safety
incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}. Some machine learning approaches have been implemented to
analyze the text in aviation incident and safety reports to predict contributing factors and topic models
\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing has also been proposed to analyze Air-Traffic communications
and condense significant features (e.g.~weather, runway, and heading info) into an XML language structure\cite{ragnarsdottir_language_2003}.
Transformer language models such as BERT and RoBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER)
tasks, translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload for pilots during
long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role identification tasks in the aviation domain
(e.g.~identifying pilot versus controller in communications); specifically, a pretrained BERT model was used and fine-tuned on problem-specific data
and compared to other models that performed well at speaker and role identification tasks in general\cite{guo_comparative_2022}.

% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc. A
majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is
used to transcribe speech
\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
and word lattices, have also been used in aviation alongside semi-supervised approaches to automatic speech recognition to leverage unlabeled data.
However, none of the developed methodologies use neural language models for their applications
\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.

% differences in language (transcripts vs literary english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the similarity (or, conversely, the difference) between literary English and aviation English. If the two domains are sufficiently different, then training the pretrained models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text domains has been shown to be possible even with data volumes significantly lower than the original data used to train the models \cite{raffel_exploring_2020}. Yadlowsky et al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different, model pretraining on out-of-domain data will negatively affect model performance.

\section{Data Sources}\label{sec:data_source}
Four corpora are used in this thesis. The datasets are primarily speech corpora intended for automatic speech recognition research in aviation communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and transcriptions, which are audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in sections \ref{sec:language_modeling} and \ref{sec:tokenizers}).

@@

\begin{table}
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ATCC Corpus Properties} \\
        \midrule
        Samples              & 9,556               \\
        Audio (hours)        & 72.48               \\
        Mean Sequence Length & 11.90               \\
        Std.~Sequence Length & 7.22                \\

@@

\begin{table}
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ATCO2 Corpus Properties} \\
        \midrule
        Samples              & 874                  \\
        Audio (hours)        & 1.10                 \\
        Mean Sequence Length & 12.28                \\
        Std.~Sequence Length & 5.78                 \\
        Total Tokens         & 10,733               \\
        Unique Tokens        & 786                  \\
        \bottomrule
    \end{tabular}
    \caption{Data statistics for the ATCO2 corpus.}
    \label{tab:atco2_stats}
\end{table}

@@

\begin{table}
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ATCOSIM Corpus Properties} \\
        \midrule
        Samples              & 9,556                  \\
        Audio (hours)        & 10.69                  \\
        Mean Sequence Length & 11.29                  \\
        Std.~Sequence Length & 4.17                   \\
        Total Tokens         & 107,881                \\
        Unique Tokens        & 827                    \\
        \bottomrule
    \end{tabular}
    \caption{Data statistics for the ATCOSIM corpus.}
    \label{tab:atcosim_stats}
\end{table}

@@

\begin{table}
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ZCU CZ ATC Corpus Properties} \\
        \midrule
        Samples              & 14,435                    \\
        Audio (hours)        & 20.58                     \\
        Mean Sequence Length & 10.05                     \\
        Std.~Sequence Length & 5.50                      \\
        Total Tokens         & 145,107                   \\
        Unique Tokens        & 3,100                     \\
        \bottomrule
    \end{tabular}
    \caption{Data statistics for ZCU CZ ATC corpus.}
    \label{tab:zcu_cz_atc_stats}
\end{table}

\section{Data Processing}\label{sec:data_processing}
All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a result, the data in these corpora all have different formats and need to be processed into a common format to be used together, with the goal being to essentially create a ``super-corpus'' that includes all the corpora listed above (i.e., a superset of the other corpora).

All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used, although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by parsing from the corpora-specific formats where necessary, e.g.~Lisp in ATCC, XML in ATCO2 and ZCU CZ ATC, and plain text in ATCOSIM, the transcriber annotations are removed where necessary\footnote{A copy of the original data is made, then the annotations are removed from the copy.}, and the resulting text is written to a file in the standard corpus format (an ASCII text file with one line in the file corresponding to one sample from the corpus).

@@

One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts, so including as much data as possible regarding spoken word realization is useful given the context. The \lstinline|[HES]| token is used as a catch-all fallback token if no knowledge about the utterance from the transcript can be extracted.
\section{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in the corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus. The same statistics were also calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models. Note that this is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.

@@

\section{Language Modeling}\label{sec:language_modeling}
% bert, roberta
% n-gram
This section introduces the chosen language model architectures and their pretrained checkpoints. Two transformer-based architectures, BERT and RoBERTa, and one statistical/machine learning-based approach: N-gram language modeling.

>>>>

\usepackage{cite}
\usepackage{listings}
\usepackage{multicol}
\usepackage{newtxtext}
\usepackage{setspace}
\usepackage{sectsty}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\usepackage{url}
\usepackage[T1]{fontenc}
% \usepackage[numbib]{tocbibind}
\usepackage[lmargin=1.25in,rmargin=1.25in,tmargin=1.0in,bmargin=1.0in]{geometry}

@@

\section{Introduction}\label{sec:introduction}
There have been noticeable and significant advancements in the general domain of artificial intelligence (AI) recently. Natural
language processing and language modeling technologies, in particular, have made tremendous advances over the past two decades
\cite{hannun_deep_2014,lee_unsupervised_2009,li_jasper_2019,kriman_quartznet_2020}, achieving performance levels good enough to be brought to
consumer-facing products such as personal assistants, search engines, phones, televisions, etc. Due to the success in the general domain, applications
have begun to spin off into problem-specific domains, such as (and most notably for this work) aviation. For example, several speech recognition
systems with rule- and machine learning-based natural language processing, inference, and semantic extraction algorithms have been implemented at
EUROCONTROL simulation centers in Europe to train air traffic controllers and study the potential effects of AI-augmented air traffic management (ATM)
systems on controller workloads. Preliminary studies have found that AI-augmented ATM systems reduce controller workloads and fuel consumption of
aircraft by optimizing aircraft routes in the airspace and reducing departure/arrival times, thereby reducing fuel consumption
\cite{helmke_quantifying_2017}.

At the time of this writing, the current focus of general machine learning and deep learning algorithms in the language modeling domains has been on a
variety of deep learning methods/architectures, namely, recurrent, convolutional, and transformer neural networks, due to the availability of large
natural language corpora such as WikiText \cite{merity_pointer_2016} and IMDB \cite{maas_learning_2011}. In contrast to the general domain, the
aviation domain still prefers and sees much success with traditional machine learning models for language modeling applications. Typically, the most
prevalent of the traditional models are N-grams, word lattices, and rule-based approaches, depending on the application and operating environment of
the models. The most likely reason for this divergence between the domains is the difference and relative lack of labeled data in the aviation domain.
By comparison, the largest corpora in the general domain are either WikiText or the BookCorpus, which contain trillions of tokens, whereas, for
aviation, the largest corpus is Air Traffic Control Complete with approximately 26 hours of labeled data amounting to just over 300,000
tokens\footnote{Air Traffic Control Complete has approximately 70 hours of audio data, however, about 26 hours of the 70 are labeled.}. This disparity
in data availability is exacerbated by the fact that aviation English uses a specialized vocabulary in addition to specialized pronunciation for some
words to enhance understandability in a radiotelephonic (R/T) medium \cite{paltridge_handbook_2013}. This makes the data labeling process
significantly more difficult because it not only requires two types of data (audio and text) to create labels, but also requires labelers to have
domain knowledge and experience to create effective transcriptions. This, in addition to the following facets, further contributes to the difficulty
of transcription: (1) the medium of communication, (2) the environment in which the communication occurs\footnote{For example, a noisy and possibly
    low fidelity medium, depending on the equipment used to transmit and receive communications}, (3) a noisy and active communication environment
i.e.~an active flight deck under the influence of wind, engine noise, etc., (4) an active Air Traffic Control (ATC) tower and channels, and (5) a
rapid rate of speech. Pilots are also operating sophisticated equipment during while communicating, requiring significant attention and mental
resources further contributing to the complexity of the communications \cite{paltridge_handbook_2013}.

% connect complexity of aviation english to utility/necessity of NLP models (ASR and LM) in the aviation domain
The potential for speech recognition and language modeling algorithms to reach better than human performance (for example, the Stanford Question
Answering Dataset (SQuAD) benchmark has recorded results consistently better than humans since
2019\footnote{\url{https://rajpurkar.github.io/SQuAD-explorer/}}) \cite{zhang_ai_2022} suggests that these NLP algorithms could have high utility in the
aviation domain. If they can achieve better-than-human performance in the aviation domain and be used with pilot and air traffic controller systems,
they could easily boost efficiency, reduce workloads, and reduce and mitigate errors during normal operations.

The methods and results laid out in this thesis were done with the intention of creating a streamlined process for training and testing deep
learning-based NLP methods, specifically language modeling. The results from the language models suggest that the volume of text used to train the
language models is not yet sufficient to produce results akin to those in the conversational and/or literary English domains. Additionally, the
results in Section \ref{sec:pretraining_results} show that language models trained ``from scratch'' on in-domain data perform better than models
trained on out-of-domain data and transfer learned on in-domain data.

\subsection{Problem Statement}\label{sec:problem_statement}
Other work has shown that the word error rates of automatic speech recognition models can and have been significantly reduced by using language models
to determine the most likely token at a time step in a sequence \cite{li_jasper_2019,kriman_quartznet_2020}. Language models have also been shown to
be effective for natural language processing tasks such as named entity recognition \cite{devlin_bert_2019,liu_roberta_2019}. Speech recognition and
callsign detection (CSD) have become popular tasks for machine learning applications in aviation
\cite{pellegrini_airbus_2019,delpech_real-life_2018,helmke_quantifying_2017}; callsign detection is a task analogous to named entity recognition since
the elements of a callsign in a transcription can be treated as named entities in a sentence. For example, in the following transcription

\begin{quote}
    skyshuttle one one four zero now descend flight level three three zero
\end{quote}

\noindent
the sequence ``skyshuttle one one four zero'' should be detected as the callsign. This is similar to detecting ``EU'', ``German'', and ``British'' as
named entities in the sequence below.

\begin{quote}
    EU rejects German call to boycott British lamb
\end{quote}

The key component missing for both of these applications is an effectively pretrained language model that can be used for these downstream tasks, so
the goal of this thesis is to run several experiments using transformer-based language models to determine the most effective strategy for pretraining
the model for downstream tasks. Recommended best practice is to use a pre-existing language model checkpoint and fine-tune on additional data
\cite{devlin_bert_2019,liu_roberta_2019,lewis_bart_2019}, although recent work suggests that if domains are sufficiently different, the performance
of the fine-tuned model will be limited \cite{yadlowsky_pretraining_2023}. The main point of comparison in this thesis will be the performance of
pretrained models fine-tuned on in-domain (aviation) data and models trained from scratch on in-domain data.

\subsection{Notation and Terminology}
This section is included for the sake of clarity throughout this thesis. In pursuit of that, a few terms and notations are defined below.

@@

\noindent
as the union of the set $A$ and a set containing the element $x$. The equation below can be read as adding the element $x$ to the set $A$:

\begin{equation*}
    A \leftarrow A; x
\end{equation*}

The following words (in bold) are used throughout the thesis as they relate to language modeling and are defined concretely as follows:

The definition of words and sentences stem from their definitions in the English language, succinctly, as defined in \cite{noauthor_american_2011}:

\begin{itemize}
    \item A \textbf{word} is the representation in writing of a sound or combination of sounds that symbolizes and communicates a meaning
    \item A \textbf{sentence} is a grammatical unit that is syntactically independent that is expressed or understood and contains at least one finite
          verb
\end{itemize}

\noindent
In terms of the structure of individual samples in a corpus, we will consider each sample in the corpus to be one sentence that is made up of words
that are separated by whitespace (one or more spaces) and/or punctuation. Structurally, we consider a word to be made up of one or more characters
as defined by the unicode standard \cite{allen_unicode_2012} and a sentence to be made up of one or more words.

\begin{itemize}
    \item A \textbf{token} shall be defined as the smallest, fundamental component of a sequence; these can be single characters, groups of
          characters, punctuation, etc. \cite{devlin_bert_2019,wu_googles_2016,sennrich_neural_2016}
    \item A \textbf{sequence} shall be defined as a set of tokens (distinct from a sentence)
\end{itemize}

\noindent
As mentioned above, we consider each sample in a corpus to be a sentence, then as the corpus is collated, processed, and tokenized, each sample will
be processed into sequences of tokens. These sequences are what will be analyzed and modeled by the language models.

The rules by which words in sentences are broken up into tokens are defined by tokenization algorithms, otherwise known as tokenizers (elaborated
further in Section \ref{sec:tokenizers}). While tokens can be individual characters or words as they appear in the original sample, this is not always
the case. The intention of most tokenization algorithms is the flexibility to break up unknown words into smaller tokens known to the tokenizer
\cite{wu_googles_2016,sennrich_neural_2016}.

\section{Background \& Literature Review}\label{sec:lit_review}

\subsection{Transformer Neural Networks}
% from thesis proposal; revise and add to this as more references are used
% transformer history (background)
The transformer neural network architecture was proposed in 2017 for Neural Machine Translation tasks and immediately achieved state-of-the-art
performance on language translation tasks \cite{vaswani_attention_2017}. Transformer architectures are extremely effective at learning representations
and understandings of languages to predict token probabilities instead of transforming one language into another
\cite{devlin_bert_2019,liu_roberta_2019}. Transformers have also been found to be very effective at other NLP tasks such as prompt completion
and sentiment analysis among others (i.e.~auto-regressive and sequence classification tasks, respectively)
\cite{lewis_bart_2019,radford_improving_2018}. Devlin et al.~\cite{devlin_bert_2019} proposed pretraining methods such as masked language modeling
and next sentence prediction to improve natural language understanding for downstream tasks, which was further expanded upon by Liu et
al.~\cite{liu_roberta_2019} by training longer, with larger batch sizes, and removing the next sentence prediction task.

\subsection{Tokenizer Algorithms}
% tokenizer algorithms
In language modeling and natural language processing tasks, the out-of-vocabulary (OOV) problem is prevalent and has been addressed by the
introduction of tokenizer algorithms. WordPiece and Byte Pair Encoding (BPE) have been used to solve OOV problems while simultaneously minimizing
vocabulary sizes and maximizing the likelihood of tokens in sequences \cite{wu_googles_2016,schuster_japanese_2012,sennrich_neural_2016}.

\subsection{Language Modeling \& Natural Language Processing in Aviation}
% LMs and NLP in aero. domain
Various natural language processing methods have been applied to the aviation domain to help deal with miscommunications and try to mitigate safety
incidents\cite{ragnarsdottir_language_2003,tanguy_natural_2016,madeira_machine_2021}. Some machine learning approaches have been implemented to
analyze the text in aviation incident and safety reports to predict contributing factors and topic models
\cite{tanguy_natural_2016,madeira_machine_2021}. An ASR system with NLP post-processing has also been proposed to analyze Air-Traffic communications
and condense significant features (e.g.~weather, runway, and heading info) into an XML language structure\cite{ragnarsdottir_language_2003}.
Transformer language models such as BERT and RoBERTa have been applied to notice to airmen (NOTAM) messages to perform named entity recognition (NER)
tasks, translation tasks (between notations; e.g.~NOTAM notation to Airlang to make parsing tasks easier) and reduce the workload for pilots during
long-haul flights\cite{arnold_knowledge_2022}. Transformer models have also been used for speaker role identification tasks in the aviation domain
(e.g.~identifying pilot versus controller in communications); specifically, a pretrained BERT model was used and fine-tuned on problem-specific data
and compared to other models that performed well at speaker and role identification tasks in general\cite{guo_comparative_2022}.

\subsection{Automatic Speech Recognition in Aviation}
% ASR context (suggest some need for semi-supervised approaches and therefore LMs)
Automatic speech recognition (ASR) is seeing increased use in aviation for tasks such as transcription, callsign detection, speaker identification, etc. A
majority of these approaches use machine learning approaches (as opposed to deep learning) where acoustic, pronunciation, and language modeling is
used to transcribe speech
\cite{guo_comparative_2022,smidl_air_2019,zuluaga-gomez_automatic_2020,badrinath_automatic_2022,hofbauer_atcosim_2008,helmke_quantifying_2017}.
Language models have been used in the general domain (both neural and machine learning-based approaches) to boost the performance of speech
recognition models \cite{han_contextnet_2020,kriman_quartznet_2020,majumdar_citrinet_2021}. Machine learning-based language models, usually N-grams
and word lattices, have also been used in aviation alongside semi-supervised approaches to automatic speech recognition to leverage unlabeled data.
However, none of the developed methodologies use neural language models for their applications
\cite{zuluaga-gomez_contextual_2021,srinivasamurthy_semi-supervised_2017,badrinath_automatic_2022}.

ASR and speech synthesis technologies have seen use for improving communications and reducing pilot workload as early as 1981
\cite{money_aural_1981,north_application_1984,north_systems_1984,mayer_electronic_1981}. Early approaches focused on developing and improving speech
synthesis technologies to shift pilot focus in military and general aviation cockpits to urgent systems or alarms
\cite{mayer_electronic_1981,money_aural_1981,mosko_clear_1981}. ASR system concepts and technologies were also initially developed to reduce pilot
workloads and automate simple tasks; these systems were somewhat rigid with fixed vocabularies and keywords, but were able to successfully reduce
workload and received positive feedback from pilots \cite{north_application_1984,cotton_development_1983}. Similar ASR technologies were later applied
to Air Traffic Control (ATC) contexts to reduce controller workloads and increase the efficiency of Air Traffic Management (ATM) solutions
\cite{schafer_context-sensitive_2000,lechner_voice_2002,ragnarsdottir_language_2003,cordero_automated_2012}. As ASR technologies have become more
sophisticated and more aviation-specific speech corpora have become available
\cite{godfrey_air_1994,smidl_air_2019,hofbauer_atcosim_2008,szoke_detecting_2021,segura_hiwire_2007,graglia_vocalise_2005,delpech_real-life_2018,zuluaga-gomez_atco2_2023}
there have been more applications of ASR to aviation
\cite{moere_implementing_2009,gurluk_assistant_2015,helmke_assistant-based_2015,helmke_machine_2020,helmke_readback_2021,kleinert_automated_2021,badrinath_automatic_2022,shi_end--end_2022,xiao_speech_2022,zuluaga-gomez_atco2_2023}.

\subsection{Differences in Language}
% differences in language (aviation vs casual english), narrow model selection
The use of the English language in aviation has been shown to be sufficiently different enough to be classified under its own category (aviation
English) \cite{paltridge_handbook_2013}. The issue with using pretrained checkpoints of transformer-based language models is characterizing the
similarity (or, conversely, the difference) between casual and technical English i.e., in this case, the differences between written English, as in
books and articles, and the spoken, transcribed English used in aviation. If the two domains are sufficiently different, then training the pretrained
models on aviation English data would amount to transfer learning between specific domains. Transfer learning between text domains has been shown to
be possible even with data volumes significantly lower than the original data used to train the models \cite{raffel_exploring_2020}. Yadlowsky et
al.~\cite{yadlowsky_pretraining_2023} recently showed that if problem domains are sufficiently different, model pretraining on out-of-domain data will
negatively affect model performance.

\section{Data Sources}\label{sec:data_source}
Four corpora are used in this thesis. The datasets are primarily speech corpora intended for automatic speech recognition research in aviation
communications, linguistics research in aviation English, or both. All data in all four corpora include both recordings and transcriptions, which are
audio and text, respectively. The transcriptions are used to pretrain and train the language models and tokenizers (described in sections
\ref{sec:language_models} and \ref{sec:tokenizers}).

@@

\begin{table}[!t]
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ATCC Corpus Properties} \\
        \midrule
        Samples              & 9,556               \\
        Audio (hours)        & 72.48               \\
        Mean Sequence Length & 11.90               \\
        Std.~Sequence Length & 7.22                \\

@@

\begin{table}[!t]
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ATCO2 Corpus Properties} \\
        \midrule
        Samples              & 874                  \\
        Audio (hours)        & 1.10                 \\
        Mean Sequence Length & 12.28                \\
        Std.~Sequence Length & 5.78                 \\
        Total Tokens         & 10,733               \\
        Unique Tokens        & 786                  \\
        \bottomrule
    \end{tabular}
    \caption{Data statistics for the ATCO2 corpus.}
    \label{tab:atco2_stats}
\end{table}

@@

\begin{table}[!t]
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ATCOSIM Corpus Properties} \\
        \midrule
        Samples              & 9,556                  \\
        Audio (hours)        & 10.69                  \\
        Mean Sequence Length & 11.29                  \\
        Std.~Sequence Length & 4.17                   \\
        Total Tokens         & 107,881                \\
        Unique Tokens        & 827                    \\
        \bottomrule
    \end{tabular}
    \caption{Data statistics for the ATCOSIM corpus.}
    \label{tab:atcosim_stats}
\end{table}

@@

\begin{table}[!t]
    \centering
    \begin{tabular}{l r}
        \toprule
        \multicolumn{2}{c}{ZCU CZ ATC Corpus Properties} \\
        \midrule
        Samples              & 14,435                    \\
        Audio (hours)        & 20.58                     \\
        Mean Sequence Length & 10.05                     \\
        Std.~Sequence Length & 5.50                      \\
        Total Tokens         & 145,107                   \\
        Unique Tokens        & 3,100                     \\
        \bottomrule
    \end{tabular}
    \caption{Data statistics for ZCU CZ ATC corpus.}
    \label{tab:zcu_cz_atc_stats}
\end{table}

\section{Data Preparation}\label{sec:data_preparation}
All corpora listed above (see Section \ref{sec:data_source}) are created at different times, for various purposes, and by multiple authors. As a
result, the data in these corpora all have different formats. ATCC, for example, uses Lisp-style lists as below:

\begin{lstlisting}
((FROM NERA3788) (NUM L02F1-0001)
(TO F1-1)
(TEXT THOUSAND ONE NINETY WE (QUOTE LL) GIVE YOU THAT ON THE SPEED AND WE (QUOTE RE) CLEARED FOR THE APPROACH AH NERA THIRTY SEVEN EIGHTY EIGHT WE (QUOTE LL) HOLD SHORT OF TWO SEVEN)
(TIMES 1.49 6.57))
\end{lstlisting}

ATCO2 and ZCU CZ ATC use XML to isolate transcribed segments of speech in the audio recordings, although the format used between the two differs.
ATCO2 uses a form like below:

\clearpage
\begin{lstlisting}
<data>
<segment>
<start>3.79</start>
<end>6.85</end>
<speaker>B</speaker>
<speaker_label>OK-PMB</speaker_label>
<text>level one hundred Oscar Kilo Papa Mike Bravo</text>
<tags>
<correct>0</correct>
<correct_transcript>1</correct_transcript>
<correct_tagging>0</correct_tagging>
<non_english>0</non_english>
</tags>
</segment>
</data>
\end{lstlisting}

\clearpage
ZCU CZ ATC uses a different nesting and labeling style, like below\footnote{Some information is excluded for readability.}:

\begin{lstlisting}
<Trans audio_filename="e2_ACCU-0agmXf.wav">
<Episode>
<Section type="report" startTime="0" endTime="61.47">
<Turn startTime="0" endTime="61.47">
<Sync time="0.000"/>
[ground]Skyshuttle 1 1 4 0 now descend FL 3 3 0[speaker]
<Sync time="4.790"/>
[air]descending FL 3
<Sync time="6.650"/>
[air]K O Z
<Sync time="9.640"/>
[ground_|]Austrian 3 2 3 G climb FL 3 4 0 [|_ground][air_|]level 3 4 0 Austrian 3 2 3 G[|_air]
</Turn>
</Section>
</Episode>
</Trans>
\end{lstlisting}

Lastly, ATCOSIM uses a text-only format with transcriptions occurring on their own lines in text files:

\begin{lstlisting}
[FRAGMENT] contact geneva one two eight decimal one five good bye
\end{lstlisting}

\noindent
Each text file corresponds to the audio file from which the speech was transcribed.

Due to the differing formats above, and the intention to aggregate and use all these transcripts together, all transcripts need to be processed into
a common format to be used together.

% TODO: fix; lost some text somehow
All data was processed in Python using primarily built-in functions/modules\footnote{The \lstinline|re| and \lstinline|xml| modules were also used,
    although these are built into most Python distributions by default.}. The text data is extracted from the corpus transcripts by reading each
transcript into the main memory of the system according to the format prescribed by the corpus. The text corresponding to the transcript is isolated,
copied, and stored to a separate region of memory (a Python list of strings) that corresponds to the original corpus. The resulting isolated
transcripts are written to files that correspond to the original corpus as well as combined, shuffled, and written to a file in the standard corpus
format (an ASCII text file with one line in the file corresponding to one sample from the corpus). The resulting output is four corpus files that
correspond to the original corpora in addition to a fifth corpus file with the samples from all four corpora combined and shuffled for later use.

@@

One of the intended downstream use cases of the language models developed from this work is decoding speech recognition predictions into transcripts,
so including as much data as possible regarding spoken word realization is useful given the context. The \lstinline|[HES]| token is used as a
fallback token if no knowledge about the utterance can be extracted from the transcript. This usually occurs when the original transcript did not
include any detail regarding the special case of an utterance, for example, an \lstinline|[UNINTELLIGIBLE]| token to mark unintelligible speech, or
\lstinline|[UNK]| to mark an unknown word or utterance that could not be transcribed. The \lstinline|[HES]| token is therefore used as a way of
unifying the different approaches to mark hesitations, unintelligible speech, etc. when a special token was used with no indication of the realized
speech.

\subsection{Data Analysis}\label{sec:data_analysis}
% sequence lengths, unique tokens, total number of tokens, total number of samples, breakdown by each dataset
Relevant corpora statistics were observed or calculated and are presented in Table \ref{tab:corpora_stats}. This includes the number of samples in the
corpus, the mean sequence length per sample, the number of unique tokens, and the total number of tokens in the corpus. The same statistics were also
calculated across corpora since all four corpora will be combined for training and testing the language and speech recognition models. Note that this
is based on the raw data after loading and unifying the corpora before preprocessing for model training/testing and outlier removal.

@@

\section{Model Architectures}
